---
title: "Delaware & Lehigh Heritage Half Marathon"
author: "Drew McWilliams"
date: "January 2, 2017"
output:
    html_document: 
      keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r maincode, echo=FALSE, warning=FALSE,include=FALSE}
    require(XML)
    library(plyr)
    library(dplyr)
    library(stringr)
    library(ggmap)
    library(leaflet)
    library(plotly)
    library(tidyr)
    setwd("~/Coursera/Developing Data Products/FirstApp/DevelopingDataProducts/Assignment1")
    source('time_match.R')
    xml_file<-choose.files()
    
    xml_File<-xmlInternalTreeParse(xml_file)
    
#pull data in from tcx fil
    data<-xmlParse(xml_file)
#put into a list
    xml_data<-xmlToList(data)
#Filter the XML list down to get to the workout data
    temps <- xml_data[["Activities"]][["Activity"]][["Lap"]][["Track"]]
#Transfer content into a data frame
    df2<-ldply(temps, data.frame)
      
    colnames(df2)<-c("Label","Time","Latitude","Longitude","Meters_Altitude","Meters_Distance")
 
# Convert columns to numeric    
    df2$Meters_Distance<- as.numeric(as.character(df2$Meters_Distance))
    df2$Meters_Altitude<- as.double(as.character(df2$Meters_Altitude))
    df2$Time<- str_replace_all(str_replace_all(str_replace_all(as.character(df2$Time), "T", " "),"\n",""),"\t","")
    df2$Time<- as.POSIXct(substr(df2$Time, 1,19), format="%Y-%m-%d %H:%M:%S")
    df2$Latitude<- as.numeric(as.character(df2$Latitude))
    df2$Longitude<- as.numeric(as.character(df2$Longitude))
#append extra columns
    df2<-mutate(df2, Meters_Interval=Meters_Distance- lag(Meters_Distance, default=0))
    df2<-mutate(df2, Miles=floor(1+(Meters_Distance/1609.344)))
    df2<-mutate(df2, Elapsed_Distance=Meters_Distance/1609.344)
    df2<-mutate(df2,DateTime=as.POSIXct(Time))
    df2<-mutate(df2,Elapsed_Time=(DateTime-DateTime[1])/60)
    df2<-mutate(df2, RunSpeed_MPH=(Meters_Interval*3600)/1609.344)
    df2<-mutate(df2, speed_delta=RunSpeed_MPH-lag(RunSpeed_MPH, default=0))
    df2$min<-format(df2$DateTime, "%H:%M")
    df2$Feet_Altitude<-df2$Meters_Altitude*3.28084
    df2$Speed_color <- ifelse(df2$RunSpeed_MPH >=lag(df2$RunSpeed_MPH), "Green", "Yellow")
    df2$Speed_color[which(is.na(df2$Speed_color))]<-"Blue"
##Assemble Runmap
    
    
 
   ## interactive_map(mean(df2$Latitude), mean(df2$Longitude))

   center<-as.array(c(mean(df2$Longitude), mean(df2$Latitude)))

##Weather data here:
    res <- revgeocode(center, output="more")
    zCode<-res$postal_code
    wd<-format(df2[1,"DateTime"],"%Y%/%m%/%d")
    w_url<-paste("https://www.wunderground.com/history/airport/KXLL/",wd,"/DailyHistory.html?reqdb.zip=",zCode,"&reqdb.magic=1&reqdb.wmo=99999&format=1", sep="")
    weather<-read.csv(url(w_url))
##Clean up weather data
    weather$DateUTC.br...<-as.character(weather$DateUTC.br...)
    colnames(weather)<-c("TimEDT","TemperatureF","Dew.PointF","Humidity","Sea.Level.PressureIn","VisibilityMPH","Wind.Direction","Wind.SpeedMPH","Gust.SpeedMPH","PrecipitationIn","Events","Conditions","WindDirDegrees","DateUTC")
    weather$DateUTC<-gsub("<br />","",weather$DateUTC)
    weather$DateUTC<-as.POSIXlt(weather$DateUTC)
    weather$DateEST<-weather$DateUTC
    weather$DateEST$hour<-weather$DateEST$hour-4
    
##merge weather with Run data
    df2$index<-sapply(df2$DateTime,Match_Time)
    df2$TemperatureF<-weather[df2$index,2]
    df2$Dew.PointF<-weather[df2$index,3]
    df2$Humidity<-weather[df2$index,4]
    df2$Sea.Level.PressureIn<-weather[df2$index,5]
    df2$VisibilityMPH<-weather[df2$index,6]
    df2$Wind.Direction<-weather[df2$index,7]
    df2$Wind.SpeedMPH<-weather[df2$index,8]
    df2$Gust.SpeedMPH<-weather[df2$index,9]
    df2$PrecipitationIn<-weather[df2$index,10]
    df2$Events<-weather[df2$index,11]
    df2$Conditions<-weather[df2$index,12]
    df2$WindDirDegrees<-weather[df2$index,13]
```

## Overview

The Delaware & Lehigh Heritage Half Marathon (D&L Half Marathon) was run on November 6th, 2016.

As a participant, I wore my smartphone that ran an app to track the event.  I was able to extract this data from the application in the form of a .tcx file (Training Center XML).

*Quick Note on Training Center XML

tcx files  are XML-formatted files that were designed for GPS tracking.  Garmin initially developed these and they are presently used by most all Activity Tracking applications.  Most applications, including the one that I used captures information every second that it is tracking an activity.

## Data sets
In this analysis, I used data that I pulled from the tcx file about my run and combined it with data from Weather Underground for that date, that was observed on a 15 minutes interval at a weather station from a nearby airport.  Some features from these Data Sets.

Activity Data

- Date/Time 
- Position (Lat/Long)
- Altitude
- Distance

Weather Data
- Time
- Temperature(F)
- Humidity
- Precipitation
- Wind (speed, direction, gust)

After Merging the data sets and creating additional dervied columns, the result was the following:

```{r Summary, echo=FALSE}
summary(df2)

```

## Plot of Avg Speed

The plot below displays Averge speed, change in Altitude, change in Temperature, Change in Humidity for each minute of the run.

```{r pressure, echo=FALSE, warning=FALSE, message=FALSE}
    avg_Grp<-as.data.frame(df2%>% group_by(min)%>% 
                               summarize(avg_Speed=mean(RunSpeed_MPH), avg_temp=mean(TemperatureF), avg_alt=mean(Feet_Altitude),avg_humid=mean(Humidity)))
    def_temp<-avg_Grp[1,"avg_temp"]
    def_alt<-avg_Grp[1,"avg_alt"]
    def_hum<-avg_Grp[1,"avg_humid"]
    avg_Grp<-avg_Grp%>% mutate(Temp_change=avg_temp-lag(avg_temp, default=def_temp))%>%
        mutate(Alt_change=avg_alt-lag(avg_alt, default=def_alt))%>%
        mutate(Humid_change=avg_humid-lag(avg_humid, default=def_hum))
    plot_Df<-avg_Grp[,c(1,2,6,7,8)]
    plot_Df<-gather(plot_Df, measurement, value, 2:5)
    plot_ly(plot_Df, x = plot_Df$min, y = plot_Df$value,  color = plot_Df$measurement)
```

## Course Map
The map displayed below is the course of the run as recorded that day from the app on my phone.  The start of the race is denoted by a Marker in the most Northern Part of the course, with an addtional marker for the finish to the south.
Addtional markers have been included to mark the highest point of the course and the point of my fastest speed.
```{r map, echo=FALSE}
interactive_map(mean(df2$Latitude), mean(df2$Longitude))

```


Code and RMarkdown for this doc available:
https://github.com/drewmcwilly/DevelopingDataProducts/tree/master/Assignment1

